---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load data.
```{r load, echo=TRUE}
ActivityData <- read.csv("activity.csv")
```

Convert date column to date class.
```{r date, echo=TRUE}
ActivityData$date <- as.Date(as.character(ActivityData$date), "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
Calculate the total number of steps per day.
```{r steps_per_day, echo=TRUE, results='asis'}
DaySum <- with(ActivityData, tapply(steps, date, sum))
```

Plot a histogram of the total number of steps taken each day.
```{r histogram1, echo=TRUE}
hist(DaySum, main = "Histogram of the total number of steps taken each day", xlab = "Total number of steps taken each day")
```

Calculate the mean and the median of the total number of steps taken per day
```{r mean_median, echo=TRUE}
mean(DaySum, na.rm = TRUE)
median(DaySum, na.rm = TRUE)
```

## What is the average daily activity pattern?
A time series plot of the 5-minute interval (x-axis) and the average number of steps taken in that interval averaged across all days (y-axis)
```{r steps_per_interval, echo=TRUE}
IntSum <- with(ActivityData, tapply(steps, interval, mean, na.rm = TRUE))
plot(IntSum, type = "l", xlab = "Interval", ylab = "Number of steps", main = "Average daily activity pattern")
```

Determine the 5 minute interval with the highest number of average steps.
```{r MaxSteps, echo=TRUE}
IntSum[which.max(IntSum)]
```

Activity peaks in the morning (8:30am).

## Imputing missing values
Determine how many missing values there are in the data.
```{r NA, echo=TRUE}
colSums(is.na(ActivityData))
```

Replace the NAs with the median number of steps for that 5 minute interval.
```{r replaceNA, echo=TRUE}
library(plyr)
stepdf <- ddply(ActivityData, .(interval), summarize, median = median(steps, na.rm = TRUE))
ActivityData2 <- merge(ActivityData, stepdf, by = "interval")
for(i in 1:nrow(ActivityData2)){
  if(is.na(ActivityData2[i,2])){
    ActivityData2[i,2] <- ActivityData2[i,4] 
  }
}
```

Calculate the total number of steps per day and plot a histogram of the total number of steps taken each day.
```{r steps_per_day2, echo=TRUE, results='asis'}
ActSum2 <- with(ActivityData2, tapply(steps, date, sum))
hist(ActSum2, main = "Histogram of the total number of steps taken each day", xlab = "Total number of steps taken each day")
mean(ActSum2, na.rm = TRUE)
median(ActSum2, na.rm = TRUE)
```

Imputing the missing values lowers the mean and median values for the total number of steps take each day.

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable to show whether a date is a weekday or a weekend.
```{r day_factor, echo=TRUE}
library(dplyr)
ActivityData3 <- mutate(ActivityData2, day = ifelse(grepl("Saturday|Sunday", weekdays(date)), "weekend", "weekday"))
ActivityData3$day <- as.factor(ActivityData3$day)
```

A panel plot consisting of time series plots of the 5-minute interval (x-axis) and the average number of steps taken across all weekday days or weekend days (y-axis). 
```{r panel_plot, echo=TRUE}
sumday <- ddply(ActivityData3, .(interval, day), summarize, mean = mean(steps, na.rm = TRUE))
library(lattice)
xyplot(mean~interval|day, data = sumday, layout = c(1,2), type = "l", ylab = "Number of steps", xlab = "Interval")
```

Weekend activity is spread throughout the day whereas weekday activity is concentrated around peak in the morning.